---
title: "Iowa Prison Population Growth Analysis"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    source_code: https://github.com/kristirasmussen/iowa-prison-pop-forecasting/
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(echarts4r)
library(dplyr)
library(janitor)
library(lubridate)
# library(timetk)
library(shiny)
library(plotly)

load("./data/arima_with_seasonality_after_2012.rda") 
monthly_prison_pop <- df_new
```

Dashboard {data-icon="ion-stats-bars"}
=====================================

Column {.sidebar data-width=250}
-----------------------------------------------------------------------

```{r}
offenseTypes <- monthly_prison_pop %>%
  select(offesnse_category) %>%
  distinct() #%>%
  # filter(!offesnse_category %in% c("None","No Charge", NA)) 

selectInput("offenseType", label = h4("Offense Type"), 
    choices = sort(offenseTypes$offesnse_category),
    selected = 'A Felony-Violent',
    multiple = F)

numericInput("popAdjustPct", 
             label = h4("Population Adjustment Percent"), 
             value = 0)

filterOffenseType <- reactive(input$offenseType)
filterPopAdjust <- reactive(input$popAdjustPct)

currPop <- reactive(monthly_prison_pop %>% 
                      ungroup() %>% 
                      filter(key == 'actual') %>% 
                      filter(index == max(index)) %>% 
                      filter(offesnse_category %in% filterOffenseType()) %>% 
                      dplyr::group_by(index) %>% 
                      dplyr::summarise(count = sum(count))
                    )

proj3YrPop <- reactive(monthly_prison_pop %>% 
                          ungroup() %>% 
                          filter(key == 'forecast') %>% 
                          filter(index == floor_date(Sys.Date(),'month') + months(36)) %>% 
                          filter(offesnse_category %in% filterOffenseType()) %>% 
                          dplyr::group_by(index) %>% 
                          dplyr::summarise(count = sum(count))
                    )

proj10YrPop <- reactive(monthly_prison_pop %>% 
                          ungroup() %>% 
                          filter(key == 'forecast') %>% 
                          filter(index == floor_date(Sys.Date(),'month') + months(120)) %>%
                          filter(offesnse_category %in% filterOffenseType()) %>% 
                          dplyr::group_by(index) %>% 
                          dplyr::summarise(count = sum(count))  
                    )

```

Column {data-width=450, data-height=150}
-----------------------------------------------------------------------
### Current Prison Population

```{r}



renderValueBox({
  
  valueBox(
    value = format(round(currPop()$count,0), nsmall=0, big.mark=","),
    icon = "fa-users",
    color = "primary"
  )
})

# renderValueBox({
#   rate <- formatC(downloadRate(), digits = 1, format = "f")
#   valueBox(
#     value = rate,
#     icon = "fa-area-chart",
#     color = if (rate >= input$rateThreshold) "warning" else "primary"
#   )
# })
```

### Projected +3 Year Prison Population

```{r}

renderValueBox({
  valueBox(
    value = format(round(proj3YrPop()$count,0), nsmall=0, big.mark=","),
    icon = "fa-chart-bar",
    color = if(proj3YrPop()$count > currPop()$count) "warning" else "primary"
  )
})

```

### Projected +10 Year Prison Population

```{r}



renderValueBox({
  valueBox(
    value = format(round(proj10YrPop()$count,0), nsmall=0, big.mark=","),
    icon = "fa-chart-bar",
    color = if(proj10YrPop()$count > currPop()$count) "warning" else "primary"
  )
})

```

Column {data-width=350}
-----------------------------------------------------------------------

### 


```{r}

renderPlotly({

  fig <- monthly_prison_pop %>%
    filter(index > "2012-01-01") %>%
    filter(offesnse_category %in% filterOffenseType()) %>%
    arrange(index) %>%
    plot_ly(x = ~index, y = ~ count, type = "scatter", mode='lines', hoverinfo = "x+y") %>%
    layout(xaxis = list(title = "Date"), 
           yaxis = list(title = "Count of Incarcerated Individuals", rangemode = "tozero")
           ,title = "All Data Predictions - Aggravated Misdemeanor - Other")
  
  fig <- fig %>% add_trace(y = ~hi.95, mode = 'lines') 
  fig <- fig %>% add_trace(y = ~lo.95, mode = 'lines')
    
  fig

})

```

About {data-icon="ion-information-circled"}
=====================================

***

### About the Project

This interactive application uses publicly available data from [https://data.iowa.gov/](https://data.iowa.gov/) to forecast Iowa Prison Populations.  

Contributors:  
William Ebert, Greyson Henderson, Shivani Kharbanda, Kristi Rasmussen, Grant Ruedy, Husam Yassin

All source code can be found [here](https://github.com/kristirasmussen/iowa-prison-pop-forecasting) on GitHub.

*Note: This project was completed for the Georgia Tech Data & Visual Analytics Fall 2021 course.*
