---
title: "Iowa Prison Population Growth Analysis"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    source_code: https://github.com/kristirasmussen/iowa-prison-pop-forecasting/
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(echarts4r)
library(dplyr)
library(janitor)
library(lubridate)
library(tidyverse)
# library(timetk)
library(shiny)
library(plotly)

# Historical Exploded Data
load("./data/monthly_prison_pop.rda")

# Forecast Data
load("./data/arima_with_seasonality_after_2012.rda") 
forecast_prison_pop <- df_new

# Identify first forecast data for Adjustment Start Date filtering functionality
firstForecastDate <- forecast_prison_pop %>%
                        filter(key == 'forecast') %>%
                        dplyr::summarise(startDate = min(index))
```

Dashboard {data-icon="ion-stats-bars"}
=====================================

Column {.sidebar data-width=250}
-----------------------------------------------------------------------

```{r}
offenseTypes <- forecast_prison_pop %>%
  select(offesnse_category) %>%
  distinct() #%>%
  # filter(!offesnse_category %in% c("None","No Charge", NA)) 

selectInput("offenseType", label = h4("Offense Type"), 
    choices = sort(offenseTypes$offesnse_category),
    selected = 'A Felony-Violent',
    multiple = F)

numericInput("popAdjustPct", 
             label = h4("Population Adjustment Percent (percent change in (+/- X.XX%) format)"), 
             value = 0)

dateInput("popAdjustDateStart", 
             label = h4("Population Adjustment Effective Date"), 
             value = as.Date('2021-12-01'))

filterOffenseType <- reactive(input$offenseType)
#Pop adjustment input value - default to 0 if empty
filterPopAdjust <- reactive(ifelse(is.na(input$popAdjustPct),0,input$popAdjustPct))
#Convert adjustment value to decimal format, add 1 (e.g., +5% -> 1.05)
filterPopAdjustPct <- reactive(1 + as.double(filterPopAdjust())/100)
#Adjustment value effective date
filterPopAdjustDateStart <- reactive(as.Date(input$popAdjustDateStart))

currPop <- reactive(forecast_prison_pop %>% 
                      ungroup() %>% 
                      filter(key == 'actual') %>% 
                      filter(index == max(index)) %>% 
                      filter(offesnse_category %in% filterOffenseType()) %>% 
                      dplyr::group_by(index) %>% 
                      dplyr::summarise(count = sum(count))
                    )

proj3YrPop <- reactive(forecast_prison_pop %>% 
                          ungroup() %>% 
                          filter(key == 'forecast') %>% 
                          filter(index == floor_date(Sys.Date(),'month') + months(36)) %>% 
                          filter(offesnse_category %in% filterOffenseType()) %>% 
                          dplyr::group_by(index) %>% 
                          dplyr::mutate(count = if_else(as.Date(index)>=filterPopAdjustDateStart(), count*filterPopAdjustPct(), count)) %>% 
                          dplyr::summarise(count = pmax(sum(count),0))
                    )

proj10YrPop <- reactive(forecast_prison_pop %>% 
                          ungroup() %>% 
                          filter(key == 'forecast') %>% 
                          filter(index == floor_date(Sys.Date(),'month') + months(119)) %>%
                          filter(offesnse_category %in% filterOffenseType()) %>% 
                          dplyr::group_by(index) %>% 
                          dplyr::mutate(count = if_else(as.Date(index)>=filterPopAdjustDateStart(), count*filterPopAdjustPct(), count)) %>% 
                          dplyr::summarise(count = pmax(sum(count),0))  
                    )

```

Column {data-width=450, data-height=85}
-----------------------------------------------------------------------
### Current Prison Population

```{r}

renderValueBox({
  
  valueBox(
    value = format(round(currPop()$count,0), nsmall=0, big.mark=","),
    icon = "fa-users",
    color = "primary"
  )
})

# renderValueBox({
#   rate <- formatC(downloadRate(), digits = 1, format = "f")
#   valueBox(
#     value = rate,
#     icon = "fa-area-chart",
#     color = if (rate >= input$rateThreshold) "warning" else "primary"
#   )
# })
```

### Projected +3 Year Prison Population

```{r}

renderValueBox({
  valueBox(
    value = format(round(proj3YrPop()$count,0), nsmall=0, big.mark=","),
    icon = "fa-chart-bar",
    color = if(proj3YrPop()$count > currPop()$count) "warning" else "primary"
  )
})

```

### Projected +10 Year Prison Population

```{r}

renderValueBox({
  valueBox(
    value = format(round(proj10YrPop()$count,0), nsmall=0, big.mark=","),
    icon = "fa-chart-bar",
    color = if(proj10YrPop()$count > currPop()$count) "warning" else "primary"
  )
})

```

Column {data-width=450, data-height=90}
-----------------------------------------------------------------------
### 

```{r}
genderSplits <- monthly_prison_pop %>%
                    unite(offense_category, c(offense_classifications,offense_types), sep = "-") %>%
                    dplyr::filter(offense_category == "A Felony-Violent") %>%
                    dplyr::filter(ds >= floor_date(Sys.Date(),'month') - months(36)) %>% #limit to last 3 years??
                    dplyr::group_by(offense_category, sex) %>%
                    # now figure out how to do an average gender split over the last 3 years, or something like that
                    dplyr::summarise(total_pop = sum(n)) %>% # we'll want to divide to get the monthly amount here too??
                    pivot_wider(names_from = sex, values_from = total_pop) %>%
                    replace(is.na(.),0) %>%
                    dplyr::mutate(pct_male = Male / (Male + Female))

# renderTable(genderSplits)

renderDataTable(genderSplits, 
                escape = TRUE, 
                options = list(scrollX=FALSE, autoWidth=TRUE, paging=FALSE, info=FALSE, searching=FALSE, ordering=FALSE))

# renderDataTable(genderSplits,
#                 options = list(
#                   iDisplayLength=1,
#                   bLengthChange=0,
#                   bPaginate=0,
#                   bFilter=0,
#                   bInfo=0,
#                   bAutoWidth=1
#                 ))

```

Column {data-width=350}
-----------------------------------------------------------------------

### 


```{r}
#mutate(count = if_else(as.Date(index)>=filterPopAdjustDateStart(), count*filterPopAdjustPct(), count)) %>% 
renderPlotly({

  fig <- forecast_prison_pop %>%
    filter(index > "2012-01-01") %>%
    filter(offesnse_category %in% filterOffenseType()) %>%
    arrange(index) %>%
    mutate(count = pmax(if_else(key == "forecast" & as.Date(index)>=filterPopAdjustDateStart(), count * filterPopAdjustPct(), count),0)) %>%
    mutate(hi.95 = pmax(if_else(key == "forecast" & as.Date(index)>=filterPopAdjustDateStart(), hi.95 * filterPopAdjustPct(), hi.95),0)) %>%
    mutate(lo.95 = pmax(if_else(key == "forecast" & as.Date(index)>=filterPopAdjustDateStart(), lo.95 * filterPopAdjustPct(), lo.95),0)) %>%
    plot_ly(x = ~index, y = ~ hi.95, type = "scatter", mode='lines', line = list(color = 'transparent'), hoverinfo = "x+y", name='Upper 95% CI') %>%
    layout(xaxis = list(title = "Date"), 
           yaxis = list(title = "Count of Incarcerated Individuals", rangemode = "tozero")
           ,title = "All Data Predictions - Aggravated Misdemeanor - Other")

  fig <- fig %>% 
    add_trace(y = ~lo.95, mode = 'lines', fill = 'tonexty', fillcolor='rgba(0,100,80,0.2)', line = list(color = 'transparent'), name='Lower 95% CI') 
  
  fig <- fig %>%
    add_trace(y = ~count, mode = 'lines', line = list(color='rgb(0,100,80)'), name='Forecast')

  fig

})

```

About {data-icon="ion-information-circled"}
=====================================

***

### About the Project

This interactive application uses publicly available data from [https://data.iowa.gov/](https://data.iowa.gov/) to forecast Iowa Prison Populations.  

Contributors:  
William Ebert, Greyson Henderson, Shivani Kharbanda, Kristi Rasmussen, Grant Ruedy, Husam Yassin

All source code can be found [here](https://github.com/kristirasmussen/iowa-prison-pop-forecasting) on GitHub.

*Note: This project was completed for the Georgia Tech Data & Visual Analytics Fall 2021 course.*
